---
title: "Minibarcode analysis"
author: "Anna J Holmquist"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir =  "/Users/aholmquist/Documents/GitHub/COI-MiniBarcode/")

```

Libraries
```{r libraries, echo = F}
library(tidyverse)
library(reshape2)
library(vegan)
library(ape)
library(BAT)
```

Data
```{r csv}
# Sequences following DADA2, LULU, decontam, and selection of spider sequences 
lco_total <- read_csv("data/lco_df.csv")

# Final data set following with all clusters defined and a single sequence 
# per adult sample, produced by data_preperation code
final_df_clustered <- read.csv("data/final_df_clustered.csv")

# Site metadata
site_data <- read_csv("data/site_metadata.csv")

```

Sequencing Results
```{r}
# Number of adults, and total spiders
num_adults <- 
  lco_total %>%
  filter(!grepl("juv", sample)) %>%
  summarise(n_distinct(sample)) %>%
  pull()

print(paste0("Number of adults: ", num_adults))
# The value 1,444 was obtained from counting specimens while plating
print("Number of juveniles: 1444")
print(paste0("Number of total spiders: ", num_adults+1444))

# Number of reads
reads <- lco_total %>%
  summarise(sum(count)) %>%
  pull()
print(paste0("Number of reads: ", reads))

# Number of ASVs, OTUs in total
asv_otu <- lco_total %>%
  summarise(asv = n_distinct(asv),
            otu = n_distinct(otu)) 

print(paste0("Number of ASVs: ", pull(asv_otu[1])))
print(paste0("Number of OTUs: ", pull(asv_otu[2])))

# Number of OTUs per juvenile pool
juv <-
  lco_total %>%
  filter(grepl("juv", sample)) %>%
  group_by(sample) %>%
  summarise(otu = n_distinct(otu)) %>%
  ungroup() %>%
  summarise(mean = mean(otu),
             med = median(otu))

print(paste0("A median of ", round(juv[[1]]), " sOTUs and mean of ", 
      juv[[2]]," sOTUs for juvenile samples"))

# Number of OTUs per adult 
adults <- 
  lco_total %>%
  filter(!grepl("juv", sample)) %>%
  group_by(sample) %>%
  summarise(otu = n_distinct(otu)) %>%
  ungroup() %>%
  summarise(mean(otu),
            median(otu))

print(paste0("A median of ", round(adults[[1]]), " sOTUs and mean of ", 
      adults[[2]]," sOTUs for adult samples"))

# Number of unique OTUs in juveniles 
juv_otu <-
  lco_total %>%
  filter(grepl("juv", sample)) %>%
  distinct(otu) %>%
  pull()

adult_otu <-
  lco_total %>%
  filter(!grepl("juv", sample)) %>%
  distinct(otu) %>%
  pull()

print(paste0("Number of OTUs found only in juvenile pools: ", 
             length(juv_otu[!juv_otu %in% adult_otu])))

print(paste0("Number of OTUs found only in adult pools: ", 
             length(adult_otu[!adult_otu %in% juv_otu])))
```

Alpha diversity - summary of diversity
```{r}
final_counts <- 
  final_df_clustered %>%
  filter(threshold =="sOTU") %>%
  summarise(n_distinct(cluster),
            n_distinct(asv))
print(paste0("After filtering: ", final_counts[[1]], " sOTUs and ", final_counts[[2]], " ASVs"))

shared <- 
  final_df_clustered %>%
  filter(threshold =="sOTU") %>%
  group_by(cluster) %>%
  filter(n_distinct(mountain) == 3) %>%
  ungroup() %>%
  summarise(n_distinct(cluster))
print(paste0(shared[[1]], " sOTUs shared between all mountains"))

multi <- 
  final_df_clustered %>%
  filter(threshold =="sOTU") %>%
  group_by(cluster) %>%
  filter(n_distinct(asv) > 1) %>%
  ungroup() %>%
  summarise(n_distinct(cluster))

print(paste0(multi[[1]], " sOTUs with multiple haplotypes"))

```

Test haplotypes across mountains
```{r statistical test for haplotypes}

# Chi-squared test
otu_mount_summary <- 
  final_df_clustered %>%
  filter(threshold =="sOTU") %>%
  group_by(cluster) %>%
  summarise(mountain = n_distinct(mountain),
            asv_num = n_distinct(asv)) %>%
  ungroup()
 
contingency_table <- table(otu_mount_summary$asv_num,
                           otu_mount_summary$mountain)
chi <- chisq.test(contingency_table)

print(paste0("Significant difference in number of haplotypes by mountain (", round(chi$p.value, digits = 50), ")"))

otu_mount_summary %>%
  group_by(mountain) %>%
  summarise(mean(asv_num))

```

Hill numbers
```{r}
community_matrix <- 
  final_df_clustered %>%
  filter(threshold =="sOTU") %>%
  group_by(site, cluster) %>%
  summarise(n = n_distinct(sample)) %>%
  acast(site ~ cluster, fill = 0)

# Hill numbers
hill <- 
  renyi(community_matrix, hill = TRUE, scales = c(0,1,2)) %>%
  rownames_to_column(var = "site")
colnames(hill) <- c("site", "q0", "q1", "q2")

# Hill dataframe
hill_info <- 
  site_data %>%
  right_join(hill, by = "site") %>%
  group_by(site) %>%
  filter(row_number() == 1) %>%
  ungroup() 

```

Relationship between Hill numbers and elevation
```{r Hill by elevation}
# Diversity by elevation group
hill_summary <- 
  hill_info %>%
  group_by(elev_grouping_1) %>%
  summarise(mean_q0 = round(mean(q0), digits = 4),
            mean_q1 = round(mean(q1), digits = 4),
            mean_q2 = round(mean(q2),  digits = 4))

# ANOVA
aov_elev_q0 <- aov(q0 ~ elev_grouping_1, data = hill_info)
aov_elev_q1 <- aov(q1 ~ elev_grouping_1, data = hill_info)
aov_elev_q2 <- aov(q2 ~ elev_grouping_1, data = hill_info)

TukeyHSD(aov_elev_q0)
TukeyHSD(aov_elev_q1)
TukeyHSD(aov_elev_q2)

# Linear models across elevation
lmq0 <- lm(q0 ~ elev, data = hill_info) # Adj R2 = 0.076, p = 0.114
lmq1 <- lm(q1 ~ elev, data = hill_info) # Adj R2 = 0.208, p = 0.017
lmq2 <- lm(q2 ~ elev, data = hill_info) # Adj R2 = 0.30, p = 0.004

```

Relationship between Hill numbers and mountain
```{r Hill by mountain}
# ANOVA - mountains
aov_q0 <- aov(q0 ~ mountain, data = hill_info)
summary(aov_q0)

aov_q1 <- aov(q1 ~ mountain, data = hill_info)
summary(aov_q1)

aov_q2 <- aov(q2 ~ mountain, data = hill_info)
summary(aov_q2)

TukeyHSD(aov_q0)
TukeyHSD(aov_q1)
TukeyHSD(aov_q2)

```

Calculate phylogenetic beta diversity
```{r phylogenetic beta diversity}
# ASV community matrix
asv_community_matrix <- 
  final_df_clustered %>%
  filter(threshold =="sOTU") %>%
  group_by(site, asv) %>%
  summarise(n = n_distinct(sample)) %>%
  acast(site ~ asv, fill = 0)

# Tree figure
phylo_df <- 
  final_df_clustered %>%
  group_by(asv) %>%
  filter(row_number() == 1) %>%
  ungroup() %>%
  select(asv, family, id)

tree <- read.tree("data/phylogeny.treefile")
tree <- drop.tip(tree, "asv_lco_1")

beta <- beta(comm = asv_community_matrix, tree = tree, raref = 20)
beta_mat <- as.matrix(beta$Btotal.mean)
rownames(beta_mat) <- rownames(asv_community_matrix)
colnames(beta_mat) <- rownames(asv_community_matrix)
beta_dist <- as.dist(beta_mat)

set.seed(999)
nmds <- metaMDS(beta_dist, k = 3, tidy = T)
nmds_df <- as_tibble(nmds$points)
nmds_df$site <- rownames(nmds$points) 

# PERMANOVA
beta_perm <- adonis2(beta_dist ~ mountain + elev + mountain*elev, data = site_data)  

# PERMDISP
beta_disp_mountain <- betadisper(beta_dist, site_data$mountain) 

beta_disp_elevation <- betadisper(beta_dist, site_data$elev_grouping_1)

anova(beta_disp_mountain)
anova(beta_disp_elevation)
```

Clustering data by mountains
```{r}
# List for each mountain
dk_sotu <- final_df_clustered$cluster[final_df_clustered$mountain_letter == "DK"
                                      & final_df_clustered$threshold == "sOTU"]
dk_97 <- final_df_clustered$cluster[final_df_clustered$mountain_letter == "DK"
                                      & final_df_clustered$threshold == "0.97"]
dk_95 <- final_df_clustered$cluster[final_df_clustered$mountain_letter == "DK"
                                      & final_df_clustered$threshold == "0.95"]

tpp_sotu <- final_df_clustered$cluster[final_df_clustered$mountain_letter == "TPP"
                                      & final_df_clustered$threshold == "sOTU"]
tpp_97 <- final_df_clustered$cluster[final_df_clustered$mountain_letter == "TPP"
                                      & final_df_clustered$threshold == "0.97"]
tpp_95 <- final_df_clustered$cluster[final_df_clustered$mountain_letter == "TPP"
                                      & final_df_clustered$threshold == "0.95"]

gor_sotu <- final_df_clustered$cluster[final_df_clustered$mountain_letter == "GOR"
                                      & final_df_clustered$threshold == "sOTU"]
gor_97 <- final_df_clustered$cluster[final_df_clustered$mountain_letter == "GOR"
                                      & final_df_clustered$threshold == "0.97"]
gor_95 <- final_df_clustered$cluster[final_df_clustered$mountain_letter == "GOR"
                                      & final_df_clustered$threshold == "0.95"]

# All shared
shared_all <- 
  final_df_clustered %>%
  group_by(threshold, cluster) %>%
  summarize(mnt_num = n_distinct(mountain)) %>%
  ungroup() %>%
  group_by(threshold, mnt_num) %>%
  summarise(shared = n_distinct(cluster)) %>%
  pivot_wider(names_from = mnt_num, values_from = shared)

summary <- 
  final_df_clustered %>%
  group_by(threshold) %>%
  summarise(total = n_distinct(cluster)) %>%
  right_join(shared_all, by = "threshold")

# Number by mountain, plus add number of unique - sOTUs
table_sotu_mnt <- 
  final_df_clustered %>%
  group_by(mountain, threshold) %>%
  summarise(n = n_distinct(cluster)) %>%
  ungroup() %>%
  filter(threshold == "sOTU")

dk_sotu_unique <- dk_sotu[!dk_sotu %in% tpp_sotu & !dk_sotu %in% gor_sotu]
gor_sotu_unique <- gor_sotu[!gor_sotu %in% dk_sotu & !gor_sotu %in% tpp_sotu]
tpp_sotu_unique <- tpp_sotu[!tpp_sotu %in% dk_sotu & !tpp_sotu %in% gor_sotu]
table_sotu_mnt$unique <- c(length(dk_sotu_unique), 
                         length(gor_sotu_unique), 
                         length(tpp_sotu_unique))

table_sotu_mnt <- 
  table_sotu_mnt %>%
  mutate(percent = unique/n)

# Number by mountain, plus add number of unique - 97%
table_97_mnt <- 
  final_df_clustered %>%
  group_by(mountain, threshold) %>%
  summarise(n = n_distinct(cluster)) %>%
  ungroup() %>%
  filter(threshold == 0.97)

dk_97_unique <- dk_97[!dk_97 %in% tpp_97 & !dk_97 %in% gor_97]
gor_97_unique <- gor_97[!gor_97 %in% dk_97 & !gor_97 %in% tpp_97]
tpp_97_unique <- tpp_97[!tpp_97 %in% dk_97 & !tpp_97 %in% gor_97]
table_97_mnt$unique <- c(length(dk_97_unique), 
                         length(gor_97_unique), 
                         length(tpp_97_unique))

table_97_mnt <- 
  table_97_mnt %>%
  mutate(percent = unique/n)

# Number by mountain, plus add number of unique - 95%
table_95_mnt <- 
  final_df_clustered %>%
  group_by(mountain, threshold) %>%
  summarise(n = n_distinct(cluster)) %>%
  ungroup() %>%
  filter(threshold == 0.95)

dk_95_unique <- dk_95[!dk_95 %in% tpp_95 & !dk_95 %in% gor_95]
gor_95_unique <- gor_95[!gor_95 %in% dk_95 & !gor_95 %in% tpp_95]
tpp_95_unique <- tpp_95[!tpp_95 %in% dk_95 & !tpp_95 %in% gor_95]
table_95_mnt$unique <- c(length(dk_95_unique), 
                         length(gor_95_unique), 
                         length(tpp_95_unique))

table_95_mnt <- 
  table_95_mnt %>%
  mutate(percent = unique/n)

```

Clustering data by elevation 
```{r}
threshold <- c(0.97, 0.95)
mountains <- unique(final_df_clustered$mountain)
elevations <- unique(final_df_clustered$elev_grouping_1)

unique_elevation <- 
  final_df_clustered %>%
  group_by(threshold, mountain, elev_grouping_1) %>%
  filter(row_number() == 1) %>%
  select(threshold, mountain, elev_grouping_1) %>%
  mutate(unique = NA)

for (i in 1:length(unique_elevation$threshold)){
  
      other <- 
        final_df_clustered %>%
        filter(threshold == unique_elevation$threshold[i] & 
                 mountain != unique_elevation$mountain[i] & 
                 elev_grouping_1!= unique_elevation$elev_grouping_1[i]) %>% 
        distinct(cluster) %>%
        pull()
      
      target <- 
        final_df_clustered %>%
        filter(threshold == unique_elevation$threshold[i] & 
                 mountain == unique_elevation$mountain[i] & 
                 elev_grouping_1 == unique_elevation$elev_grouping_1[i]) %>% 
        filter(!cluster %in% other) %>%
        distinct(cluster) %>%
        pull()
      
      unique_elevation$unique[i] <- length(target)
}

unique_elevation <- 
  unique_elevation %>%
  pivot_wider(names_from = mountain, values_from = unique)

# Number in total
cluster_elev <- 
  final_df_clustered %>%
  group_by(threshold, mountain) %>%
  mutate(total_cluster = n_distinct(cluster)) %>%
  ungroup() %>%
  group_by(threshold, mountain, elev_grouping_1) %>%
  mutate(elev_cluster = n_distinct(cluster)) %>% 
  ungroup() %>%
  mutate(perc = elev_cluster / total_cluster) %>%
  group_by(threshold, mountain, elev_grouping_1) %>%
  filter(row_number() == 1) %>%
  select(threshold, mountain, elev_grouping_1,
         total_cluster, elev_cluster, perc) 

```

Matrices of shared clusters by elevation
```{r}
custom_order <- c("Ilomata: < 500m", "Ilomata: 500 - 1000m", 
                      "Ilomata: 1000 - 1500m",
                      "Dako: < 500m", "Dako: 500 - 1000m", "Dako: 1000 - 1500m",
                      "Dako: 1500 - 2000m", "Dako: >2000m",
                      "Torompupu: 500 - 1000m", "Torompupu: 1000 - 1500m",
                      "Torompupu: 1500 - 2000m", "Torompupu: >2000m")

# 97%
cluster_97 <- 
  final_df_clustered %>%
  filter(threshold == "0.97") %>%
  mutate(mnt_elev = paste0(mountain,": ", elev_grouping_1)) %>%
  distinct(mnt_elev, cluster)

mnt_elev <- unique(cluster_97$mnt_elev)
num <- length(mnt_elev)
shared_mat <- matrix(0, nrow = num, ncol = num,
                 dimnames = list(mnt_elev, mnt_elev))

for (i in 1:(num - 1)){
  for (j in (i + 1):num){
    x1 <- mnt_elev[i]
    x2 <- mnt_elev[j]
    shared <- sum(cluster_97$cluster[cluster_97$mnt_elev == x1] %in% 
                    cluster_97$cluster[cluster_97$mnt_elev == x2])
    shared_mat[x1, x2] <- shared
    shared_mat[x2, x1] <- shared
  }
}

shared_97 <- shared_mat[custom_order, custom_order]

# 95%
cluster_95 <- 
  final_df_clustered %>%
  filter(threshold == "0.95") %>%
  mutate(mnt_elev = paste0(mountain,": ", elev_grouping_1)) %>%
  distinct(mnt_elev, cluster)

mnt_elev <- unique(cluster_95$mnt_elev)
num <- length(mnt_elev)
shared_mat <- matrix(0, nrow = num, ncol = num,
                 dimnames = list(mnt_elev, mnt_elev))

for (i in 1:(num - 1)){
  for (j in (i + 1):num){
    x1 <- mnt_elev[i]
    x2 <- mnt_elev[j]
    shared <- sum(cluster_95$cluster[cluster_95$mnt_elev == x1] %in% 
                    cluster_95$cluster[cluster_95$mnt_elev == x2])
    shared_mat[x1, x2] <- shared
    shared_mat[x2, x1] <- shared
  }
}

shared_95 <- shared_mat[custom_order, custom_order]

```

Biodiversity summary
```{r summary}
# Create data set for just sOTUs
df_sotu <- 
  final_df_clustered %>%
  filter(threshold == "sOTU")

above_95 <- df_sotu %>%
  group_by(cluster) %>%
  filter(row_number() == 1) %>%
  ungroup() %>%
  filter(percent >= 95) %>%
  summarise(n_distinct(cluster))

print(paste0("Number of sOTUs with matches above 95%: ", 
             above_95[[1]]))

above_99 <- df_sotu %>%
  filter(percent > 99) %>%
  summarise(n_distinct(cluster))
print(paste0("Number of sOTUs with matches above 99%: ", 
             above_99[[1]]))

sp <- df_sotu %>%
  filter(percent > 99) %>%
  distinct(genus, species) %>% 
  mutate(sp = paste0(genus, " ", species)) %>%
  pull(sp)
print(c("Species:", sp))

families <- df_sotu %>%
  filter(!is.na(id)) %>%
  summarise(n_distinct(id))

print(paste0("Number of families: ", 
             families[[1]]))
```
